Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns

### 2019-08-01 ms/manuscript.Rmd
Update MS
- Check numbers using code

# Load analysis results
loadd(best_fit_div_models)
loadd(climate_ancova_results)
loadd(climate_plot)
loadd(climate_select)
loadd(combined_comm_div_plots)
loadd(combined_cwm_plots)
loadd(comm)
loadd(correlated_evo_test)
loadd(div_el_model_summaries)
loadd(div_t_test_results)
loadd(div_metrics_select)
loadd(fern_traits)
loadd(importance_heatmap)
loadd(pca_plot)
loadd(pca_results)
loadd(phylosig_binary_traits)
loadd(phylosig_cont_traits)
loadd(pic_results)
loadd(traits_with_tree)
```

@@

A fern community survey was carried out on Moorea comprising 17, \SI{10}{\meter} $\times$ \SI{10}{\meter} plots spanning an elevational gradient from \SIrange{20}{2000}{\meter}, with plots placed \circa every \SI{200}{\meter} [@Nitta2017]. Most plots were located along trails leading to the three main peaks on the island, Mt. Rotui (\SI{899}{\meter}), Mt. Mouaputa (\SI{890}{\meter}), and Mt. Tohiea (\SI{1206}{\meter}). Each plot was divided into 25, \SI{2}{\meter} $\times$ \SI{2}{\meter} subplots. Presence/absence of all sporophytes in each subplot was scored and summed to produce an abundance ranking for each species per plot from zero (not observed) to 25 (observed in all subplots). Epiphytic ferns were sampled by hand to \circa \SI{2}{\meter} on tree trunks or confirmed visually from the ground. Species were categorized as epiphytic or terrestrial based on field observations (see Trait Selection and Measurements below). The list of species in each plot was split into epiphytic and terrestrial groups and treated separately during data analysis, hereafter referred to as 'epiphytic' and 'terrestrial' communities. Vouchers were deposited at GH and UC.

@@

Temperature and RH of epiphytic and terrestrial habitats were measured using Hobo Pro v2 dataloggers with the RS3 Solar Radiation Shield (Onset Corporation, USA). A pair of dataloggers was installed for each plot, one mounted at \circa \SI{2}{\meter} on a tree ('epiphytic datalogger'), and one at ground level ('terrestrial datalogger'), except for the Mt. Tohiea \SI{1200}{\meter} plot, which only had an epiphytic datalogger. Temperature and RH were logged once every \SI{15}{\minute} during two survey periods: a preliminary survey from July 18, 2012 to July 7, 2013 at six plots, and a final survey from July 7, 2013 to July 5, 2014 at 15 plots. Two plots were not included in the environmental survey. We used the 2013–-2014 data for the analysis, with one exception: the terrestrial datalogger at the Mt. Tohiea \SI{393}{\meter} plot malfunctioned and was missing data for a substantial part of the final survey period (March 12--July 5, 2014), so we used data for the same dates from the preliminary survey period for this datalogger instead. Three other dataloggers also failed early during the final survey period, but lacked data from the preliminary period, so were excluded from analysis. Other days missing data for any datalogger during the final survey period were excluded from the dataset. The final dataset included 201 days of data for 13 epiphytic dataloggers and 13 terrestrial dataloggers ($\n = 26$). We calculated vapor pressure deficit (VPD) from relative humidity and temperature as implemented in the \package{plantecophys} R package [@Duursma2015].

@@

For comparative analyses, we used the dated phylogenetic tree of [@Nitta2017], which includes all species from Moorea and some from the neighboring island of Tahiti. We trimmed the tree to species occurring only on Moorea ($\n = 130$ \si{\species}) using the \function{drop.tip} function of the \package{ape} package [@Paradis2004] in R [@RCoreTeam2015]. We tested for phylogenetic signal in quantitative traits using two metrics that relate the observed trait values to those expected under a model of trait evolution following Brownian motion (BM), Blomberg's *K* [@Blomberg2003] and Pagel's $\lambda$ [@Pagel1999], with the \function{phylosig} function in the \package{phytools} R package [@Revell2012]. $\lambda$ is a scaling parameter that ranges from 0 (traits evolving randomly, as on a star phylogeny) to 1 (traits evolving along branches according to Brownian motion). \K describes the ratio between the amount of observed variance in traits vs. the amount of variance expected under a BM model of trait evolution: when $\K = 1$, traits are evolving according to BM; when $\K > 1$, traits are more conserved than expected under BM; when $\K < 1$, traits have less phylogenetic signal than expected under BM [@Blomberg2003]. We tested for significance in \K by comparing observed values of \K against a null distribution of \num{1000} trees with trait values shuffled randomly across the tips, and in $\lambda$ by conducting a likelihood ratio test comparing the log likelihoods of the observed value of $\lambda$ vs. $\lambda = 0$ (no phylogenetic signal). We tested for phylogenetic signal in qualitative traits using Fritz and Purvis' *D* [@Fritz2010], which is designed to test for signal in binary traits. Values of \D range from 0 under a model of trait evolution by BM, to 1 for random distribution of traits; negative values (greater trait conservation than BM) and values greater than 1 (trait overdispersion) are possible. We analyzed \D using the \function{phylo.d} function in the \package{caper} R package [@Orme2013], which also conducts a significance test by comparing the observed value of \D with distributions of simulated values produced under one of two models: random shuffling of traits on the tree, or simulation of a binary trait under a BM threshold model. To account for possible circularity in measuring the phylogenetic signal of traits that were scored based on taxonomy, we conducted the same analyses on a dataset excluding species that had been scored this way ("strict" dataset).

>>>>

# Load analysis results
loadd(best_fit_div_models)
loadd(climate_ancova_results)
loadd(climate_plot)
loadd(climate_select)
loadd(combined_comm_div_plots)
loadd(combined_cwm_plots)
loadd(comm)
loadd(correlated_evo_test)
loadd(daily_mean_climate)
loadd(div_el_model_summaries)
loadd(div_t_test_results)
loadd(div_metrics_select)
loadd(fern_traits)
loadd(importance_heatmap)
loadd(pca_plot)
loadd(pca_results)
loadd(phylosig_binary_traits)
loadd(phylosig_cont_traits)
loadd(pic_results)
loadd(species_list)
loadd(traits_with_tree)
```

@@

A fern community survey was carried out on Moorea comprising 17, \SI{10}{\meter} $\times$ \SI{10}{\meter} plots spanning an elevational gradient from \SIrange{20}{2000}{\meter}, with plots placed \circa every \SI{200}{\meter} [@Nitta2017]. Most plots were located along trails leading to the three main peaks on the island, Mt. Rotui (\SI{899}{\meter}), Mt. Mouaputa (\SI{890}{\meter}), and Mt. Tohiea (\SI{1206}{\meter}). Each plot was divided into 25, \SI{2}{\meter} $\times$ \SI{2}{\meter} subplots. Presence/absence of all sporophytes in each subplot was scored and summed to produce an abundance ranking for each species per plot from zero (not observed) to 25 (observed in all subplots). Epiphytic ferns were sampled by hand to \circa \SI{2}{\meter} on tree trunks or confirmed visually from the ground. Species were categorized as epiphytic or terrestrial based on field observations (see Trait Selection and Measurements below). The list of species in each plot was split into epiphytic and terrestrial groups and treated separately during data analysis, hereafter referred to as 'epiphytic' and 'terrestrial' communities. Vouchers were deposited at GH and UC (abbreviations follow Index Herbariorum [@Thiers2019]).

@@

```{r climate-numbers}
# Count total number of days in climate data
total_days <- daily_mean_climate %>% pull(date) %>% n_distinct()

# Check start and end dates
# - start July 7, 2013
assert_that(daily_mean_climate %>% pull(date) %>% min == "2013-07-07")
# - end July 5, 2014
assert_that(daily_mean_climate %>% pull(date) %>% max == "2014-07-05")

# Count total number of sites in climate data
total_sites <-
daily_mean_climate %>%
  select(site, habit) %>%
  unique() %>%
  count(habit)
```

Temperature and RH of epiphytic and terrestrial habitats were measured using Hobo Pro v2 dataloggers with the RS3 Solar Radiation Shield (Onset Corporation, USA). A pair of dataloggers was installed for each plot, one mounted at \circa \SI{2}{\meter} on a tree ('epiphytic datalogger'), and one at ground level ('terrestrial datalogger'), except for the Mt. Tohiea \SI{1200}{\meter} plot, which only had an epiphytic datalogger. Temperature and RH were logged once every \SI{15}{\minute} during two survey periods: a preliminary survey from July 18, 2012 to July 7, 2013 at six plots, and a final survey from July 7, 2013 to July 5, 2014 at 15 plots. Two plots were not included in the environmental survey. We used the 2013–-2014 data for the analysis, with one exception: the terrestrial datalogger at the Mt. Tohiea \SI{393}{\meter} plot malfunctioned and was missing data for a substantial part of the final survey period (March 12--July 5, 2014), so we used data for the same dates from the preliminary survey period for this datalogger instead. Three other dataloggers also failed early during the final survey period, but lacked data from the preliminary period, so were excluded from analysis. Other days missing data for any datalogger during the final survey period were excluded from the dataset. The final dataset included `r total_days` days of data for `r total_sites %>% filter(habit == "epiphytic") %>% pull(n)` epiphytic dataloggers and `r total_sites %>% filter(habit == "terrestrial") %>% pull(n)` terrestrial dataloggers. We calculated vapor pressure deficit (VPD) from relative humidity and temperature as implemented in the \package{plantecophys} R package [@Duursma2015].

@@

For comparative analyses, we used the dated phylogenetic tree of [@Nitta2017], which includes all species from Moorea and some from the neighboring island of Tahiti. We trimmed the tree to species occurring only on Moorea (*n* = `r length(species_list)` \si{\species}) using the \function{drop.tip} function of the \package{ape} package [@Paradis2004] in R [@RCoreTeam2015]. We tested for phylogenetic signal in quantitative traits using two metrics that relate the observed trait values to those expected under a model of trait evolution following Brownian motion (BM), Blomberg's *K* [@Blomberg2003] and Pagel's $\lambda$ [@Pagel1999], with the \function{phylosig} function in the \package{phytools} R package [@Revell2012]. $\lambda$ is a scaling parameter that ranges from 0 (traits evolving randomly, as on a star phylogeny) to 1 (traits evolving along branches according to Brownian motion). \K describes the ratio between the amount of observed variance in traits vs. the amount of variance expected under a BM model of trait evolution: when $\K = 1$, traits are evolving according to BM; when $\K > 1$, traits are more conserved than expected under BM; when $\K < 1$, traits have less phylogenetic signal than expected under BM [@Blomberg2003]. We tested for significance in \K by comparing observed values of \K against a null distribution of \num{1000} trees with trait values shuffled randomly across the tips, and in $\lambda$ by conducting a likelihood ratio test comparing the log likelihoods of the observed value of $\lambda$ vs. $\lambda = 0$ (no phylogenetic signal). We tested for phylogenetic signal in qualitative traits using Fritz and Purvis' *D* [@Fritz2010], which is designed to test for signal in binary traits. Values of \D range from 0 under a model of trait evolution by BM, to 1 for random distribution of traits; negative values (greater trait conservation than BM) and values greater than 1 (trait overdispersion) are possible. We analyzed \D using the \function{phylo.d} function in the \package{caper} R package [@Orme2013], which also conducts a significance test by comparing the observed value of \D with distributions of simulated values produced under one of two models: random shuffling of traits on the tree, or simulation of a binary trait under a BM threshold model. To account for possible circularity in measuring the phylogenetic signal of traits that were scored based on taxonomy, we conducted the same analyses on a dataset excluding species that had been scored this way ("strict" dataset).
